\documentclass[lettersize,journal]{IEEEtran}

\usepackage{amsmath,amssymb,amsthm,amsfonts}
\usepackage{array}
\usepackage{textcomp}
\usepackage{stfloats}
\usepackage{url}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{cite}
\usepackage[hidelinks]{hyperref}
\usepackage{tikz}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}

\newtheorem{proposition}{Proposition}
\newtheorem{observation}{Observation}

\hyphenation{op-tical net-works semi-conduc-tor IEEE-Xplore Ban-zhaf}

\begin{document}

\title{Voting Weight vs.\ Voting Power:\\
An Analysis of the Banzhaf Power Index in Weighted and Dual-Majority Voting Games}

\author{Cira~Chen, Allen~Gao, and JunHang~Wu%
\thanks{Department of Mathematics, Simon Fraser University.}%
\thanks{Manuscript prepared for MATH 302 Project Report.}}

\markboth{MATH 302 Project Report}{Chen \MakeLowercase{\textit{et al.}}: Voting Weight vs. Voting Power}

\maketitle

\begin{abstract}
Weighted voting systems assign unequal formal weights, but those weights do not necessarily match actual influence. This report uses the Banzhaf Power Index to study that difference in both a standard weighted voting game and a dual-majority voting system. We first analyze the game $[10:7,6,4,1]$ to show that unequal weights can yield equal power and that a player with positive weight can still have zero effective influence. We then model the Canadian Constitutional Amending Formula as a dual-threshold game requiring both population support and provincial breadth. The results show that coalition structure and quota design, rather than nominal size alone, determine effective power.
\end{abstract}

\begin{IEEEkeywords}
Banzhaf Power Index, weighted voting game, dual-majority voting, voting power, coalition, critical player.
\end{IEEEkeywords}

\section{Introduction}
\IEEEPARstart{I}{n} democratic institutions, corporate boardrooms, and international alliances, decision-making often relies on weighted voting systems. In these systems, participants hold unequal voting weights such as shares or population, and a motion passes only if the supporting players meet a pre-defined quota. However, a player's actual influence is rarely proportional to their nominal weight. This project uses the Banzhaf Power Index to reveal how mathematical thresholds can marginalize medium-sized participants or even render small members powerless. By analyzing these hidden dynamics, we provide a lens for evaluating the fairness of institutional designs, including the general amendment procedure for the Canadian Constitution.

The central question is: To what extent does voting weight correspond to actual power, and how do multi-dimensional rules reshape this relationship? We begin by modeling the theoretical system $[10:7,6,4,1]$ to identify critical players and demonstrate a power paradox in which unequal weights produce identical influence. We then extend the analysis to the Canadian Constitutional Amending Formula, which requires a double majority: the approval of at least seven provinces representing at least $50\%$ of the national population. By comparing a hypothetical population-only system with the actual 7/50 framework, we show mathematically how institutional rules can transform marginalized members from dummy players into pivotal decision-makers.

\section{Mathematical Framework and Methodology}
\subsection{Formal Modeling of Weighted Voting Systems}
A weighted voting game is defined as
\begin{equation}
[q:w_1,w_2,\dots,w_n],
\end{equation}
where $q$ is the quota required for a decision to pass, $w_i$ is the voting weight of player $i$, and $n$ is the total number of players. Each subset of players forms a coalition, and decisions are made according to whether a coalition meets the quota.

Formally, a weighted voting system can be modeled as a simple game $(N,v)$, where $N=\{1,2,\dots,n\}$ is the set of players and $v:2^N\to\{0,1\}$ is a characteristic function. For a standard weighted voting game,
\begin{equation}
v(S)=
\begin{cases}
1, & \text{if } \sum_{i\in S} w_i \ge q,\\
0, & \text{otherwise.}
\end{cases}
\end{equation}
This formulation highlights that voting systems are fundamentally set functions defined over the power set $2^N$, so their analysis is inherently combinatorial rather than purely arithmetic.

\subsection{Coalition Enumeration}
To analyze voting power, we consider all possible coalitions. For a system with $n$ players, there are $2^n-1$ nonempty coalitions. A coalition $S$ is called \emph{winning} if $\sum_{i\in S} w_i \ge q$; otherwise, it is \emph{losing}. Enumerating all coalitions allows us to evaluate systematically which players are influential in decision-making. The enumeration of all coalitions corresponds to evaluating $v(S)$ over the entire Boolean lattice $2^N$. This exponential structure explains why computational methods are required even for moderately sized systems.

\subsection{Identifying Critical Players}
A key concept in measuring voting power is the idea of a \emph{critical player}, also called a swing voter. A player $i$ is critical in a coalition $S$ if the coalition is winning and removing player $i$ makes it losing. Formally, player $i$ is critical if
\begin{equation}
\text{Weight}(S)\ge q \quad \text{and} \quad \text{Weight}(S\setminus\{i\})<q.
\end{equation}
To determine this, we apply the deletion test:
\begin{enumerate}
    \item For each winning coalition, remove one player at a time.
    \item Check whether the coalition becomes losing.
\end{enumerate}
Equivalently, the criticality condition can be written using the characteristic function as
\begin{equation}
v(S\cup\{i\})-v(S)=1.
\end{equation}
This shows that a player's influence can be interpreted as a marginal contribution over all subsets.

\subsection{Computation and Normalization of the Banzhaf Index}
Let $\eta_i$ denote the number of coalitions in which player $i$ is critical. The total number of swings for player $i$ can be expressed as
\begin{equation}
\eta_i=\sum_{S\subseteq N\setminus\{i\}} [v(S\cup\{i\})-v(S)].
\end{equation}
The normalized Banzhaf index of player $i$ is then defined as
\begin{equation}
\beta_i=\frac{\eta_i}{\sum_{j=1}^n \eta_j}.
\end{equation}
This normalized index represents the proportion of total influence held by player $i$. The combinatorial expression makes clear that two players with different weights may have identical power if their marginal contributions occur in the same number of coalitions.

\subsection{Computational Methodology and Algorithm Design}
Analyzing the Canadian Constitution is significantly more complex than the theoretical example above. With 10 provinces, there are $2^{10}-1=1023$ possible nonempty coalitions. To handle this computation, we developed a custom algorithm in Python.

The program generates the power set of all provinces and implements a dual-threshold validator.
\begin{itemize}
    \item \textbf{Iterative enumeration:} The script loops through every possible subset of provinces.
    \item \textbf{Constraint satisfaction:} For each subset $S$, the script evaluates two conditions: $\sum_{i\in S} \text{Population}_i \ge 50\%$ and $|S|\ge 7$.
    \item \textbf{Criticality mapping:} Once a winning coalition is identified, the script performs a deletion test by temporarily removing each member. If the removal causes the coalition to fail either threshold, that province's critical count $\eta_i$ is incremented.
    \item \textbf{Normalization:} The Banzhaf index $\beta_i$ is obtained by dividing each province's critical count by the sum of all critical counts across the federation.
\end{itemize}

The population weights used in our model are sourced from Statistics Canada and represent the demographic landscape used in constitutional negotiations, ranging from Ontario's $38.8\%$ to Prince Edward Island's $0.4\%$.

\subsection{Interpretation}
The Banzhaf index measures a player's ability to influence outcomes by being pivotal in winning coalitions. It therefore gives a more accurate representation of decision-making power than nominal voting weights alone.

\section{Analysis and Application}
\subsection{Theoretical Case: The Power Paradox}
To illustrate the nonlinear relationship between weight and power, we examine a four-player game. In this scenario, Player A has 7 votes, Player B has 6 votes, Player C has 4 votes, and Player D has 1 vote, with a quota of 10. This is the weighted voting game $[10:7,6,4,1]$.

Table~\ref{tab:coalition_analysis} enumerates all possible nonempty coalitions, their total weights, and the critical players within each winning coalition. For example, coalition $\{B,C\}$ has weight $6+4=10$ and is winning; removing either player reduces the weight below the quota, so both B and C are critical. Coalition $\{A,C\}$ has weight $7+4=11$, and removing either player again makes the coalition losing, so both are critical there as well.

\begin{table}[!t]
\caption{Coalition Analysis for the Weighted Voting Game $[10:7,6,4,1]$}
\label{tab:coalition_analysis}
\centering
\begin{tabular}{c c c c}
\toprule
\textbf{Coalition} & \textbf{Weight} & \textbf{Winning} & \textbf{Critical Players} \\
\midrule
$\{A\}$ & 7 & No & -- \\
$\{B\}$ & 6 & No & -- \\
$\{C\}$ & 4 & No & -- \\
$\{D\}$ & 1 & No & -- \\
$\{A,B\}$ & 13 & Yes & A, B \\
$\{A,C\}$ & 11 & Yes & A, C \\
$\{A,D\}$ & 8 & No & -- \\
$\{B,C\}$ & 10 & Yes & B, C \\
$\{B,D\}$ & 7 & No & -- \\
$\{C,D\}$ & 5 & No & -- \\
$\{A,B,C\}$ & 17 & Yes & None \\
$\{A,B,D\}$ & 14 & Yes & A, B \\
$\{A,C,D\}$ & 12 & Yes & A, C \\
$\{B,C,D\}$ & 11 & Yes & B, C \\
$\{A,B,C,D\}$ & 18 & Yes & None \\
\bottomrule
\end{tabular}
\end{table}

\begin{observation}[Equalization of Power]
By systematically enumerating all coalitions, Players A, B, and C each have 4 swings, so $\eta_A=\eta_B=\eta_C=4$. This yields an identical Banzhaf index of $33.3\%$ for all three players. Despite Player A having nearly double the votes of Player C, they hold exactly the same influence over the outcome.
\end{observation}

\begin{observation}[Dummy Player]
Player D holds 1 vote but has 0 swings, so $\eta_D=0$. This player is a dummy player: no matter which coalition D joins, D never turns a losing outcome into a winning one.
\end{observation}

\begin{figure}[!t]
\centering
\begin{tikzpicture}
\begin{axis}[
    ybar,
    bar width=12pt,
    width=\columnwidth,
    height=0.72\columnwidth,
    enlarge x limits=0.12,
    legend style={at={(0.5,-0.18)},anchor=north,legend columns=1,font=\footnotesize},
    ylabel={Percentage (\%)},
    symbolic x coords={Player A,Player B,Player C,Player D},
    xtick=data,
    xticklabel style={font=\footnotesize,align=center},
    nodes near coords,
    nodes near coords align={vertical},
    ymin=0,
    ymax=50,
    ymajorgrids=true,
    grid style=dashed
]
\addplot coordinates {(Player A,38.9) (Player B,33.3) (Player C,22.2) (Player D,5.6)};
\addplot coordinates {(Player A,33.3) (Player B,33.3) (Player C,33.3) (Player D,0.0)};
\legend{Nominal Weight, Banzhaf Power}
\end{axis}
\end{tikzpicture}
\caption{Comparison of nominal voting weight and Banzhaf power in the game $[10:7,6,4,1]$.}
\label{fig:paradox}
\end{figure}

\begin{proposition}[Nonlinearity of Voting Power]
The Banzhaf power index is not a linear function of voting weights. It depends on the number of pivotal occurrences across coalitions, so increasing a player's weight does not necessarily increase that player's number of swings.
\end{proposition}

\subsection{Real-World Application: The Canadian 7/50 Amending Formula}
We now elevate this logic to the Canadian Constitutional Amending Formula, which acts as a multi-threshold voting game. Unlike a simple weighted game, a winning coalition $S$ must satisfy two distinct mathematical quotas:
\begin{itemize}
    \item \textbf{Population quota:} $\sum_{i\in S} \mathrm{Pop}_i \ge 50\%$.
    \item \textbf{Provincial quota:} $|S|\ge 7$.
\end{itemize}

This system can be expressed by the characteristic function
\begin{equation}
v(S)=
\begin{cases}
1, & \text{if } \sum_{i\in S} \mathrm{Pop}_i \ge 50\% \text{ and } |S|\ge 7,\\
0, & \text{otherwise.}
\end{cases}
\end{equation}

Unlike standard weighted voting games, this formulation introduces a non-additive constraint because the outcome cannot be determined by weights alone. Instead, both the population total and the coalition size jointly determine whether the coalition is winning.

Using our Python simulation, we compared the actual 7/50 framework against a hypothetical population-only system to observe how the addition of a second threshold reshapes the provincial power landscape. Table~\ref{tab:power_comparison} summarizes the results.

\begin{table}[!t]
\caption{Comparison of Population, Population-Only Power, and 7/50 Power}
\label{tab:power_comparison}
\centering
\scriptsize
\resizebox{\columnwidth}{!}{%
\begin{tabular}{lrrrrrr}
\toprule
\textbf{Province} & \textbf{Pop. \%} & \textbf{Swings (Pop)} & \textbf{$\beta_i$ (Pop)} & \textbf{Swings (7/50)} & \textbf{$\beta_i$ (7/50)} & \textbf{Change} \\
\midrule
Ontario (ON)     & 38.8 & 394 & 0.5000 & 108 & 0.1463 & $-70.7\%$ \\
Quebec (QC)      & 22.3 & 118 & 0.1497 & 78  & 0.1057 & $-29.4\%$ \\
Brit. Col. (BC)  & 13.7 & 118 & 0.1497 & 78  & 0.1057 & $-29.4\%$ \\
Alberta (AB)     & 11.7 & 118 & 0.1497 & 78  & 0.1057 & $-29.4\%$ \\
Manitoba (MB)    & 3.6  & 10  & 0.0127 & 66  & 0.0894 & $+603.9\%$ \\
Sask. (SK)       & 3.1  & 10  & 0.0127 & 66  & 0.0894 & $+603.9\%$ \\
Nova Scotia (NS) & 2.6  & 10  & 0.0127 & 66  & 0.0894 & $+603.9\%$ \\
New Bruns. (NB)  & 2.1  & 6   & 0.0076 & 66  & 0.0894 & $+1076\%$ \\
Newfound. (NL)   & 1.4  & 2   & 0.0025 & 66  & 0.0894 & $+3476\%$ \\
Prince Ed. (PE)  & 0.4  & 2   & 0.0025 & 66  & 0.0894 & $+3476\%$ \\
\bottomrule
\end{tabular}%
}
\end{table}

The data reveals a systematic mismatch between population weight and actual voting power, leading to three major structural transformations.

\subsubsection*{Finding 1: Breakdown of Proportionality}
Under the hypothetical population-only system, one might expect the Banzhaf index to scale roughly with population size. While this relationship holds loosely for Ontario, it breaks down for the rest of the federation. Quebec's population share ($22.3\%$) is almost twice Alberta's ($11.7\%$), yet the two provinces have identical Banzhaf indices of $0.1497$. Their marginal contributions are structurally equivalent rather than proportionally determined.

This nonlinearity becomes even more pronounced among smaller provinces. Provinces with less than $4\%$ population share have disproportionately low power, while Newfoundland and Labrador and Prince Edward Island approach near-zero power under the population-only rule.

\begin{figure}[!t]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=0.98\columnwidth,
    height=0.78\columnwidth,
    xlabel={Population Share (\%)},
    ylabel={Banzhaf Power (\%)},
    xmin=0,
    xmax=45,
    ymin=0,
    ymax=60,
    grid=both,
    grid style={dashed},
    legend pos=north west,
    tick label style={font=\scriptsize},
    label style={font=\scriptsize},
    legend style={font=\scriptsize}
]
\addplot[domain=0:45,thick,dashed] {x};
\addlegendentry{Ideal $y=x$}
\addplot[only marks,mark=*,mark size=2.5pt] coordinates {
    (38.8,50.00)
    (22.3,14.97)
    (13.7,14.97)
    (11.7,14.97)
    (3.6,1.27)
    (3.1,1.27)
    (2.6,1.27)
    (2.1,0.76)
    (1.4,0.25)
    (0.4,0.25)
};
\addlegendentry{Provinces}
\node[anchor=south west,font=\scriptsize] at (axis cs:38.8,50.00) {ON};
\node[anchor=west,font=\scriptsize] at (axis cs:22.8,14.97) {QC};
\node[anchor=south,font=\scriptsize] at (axis cs:13.7,16.00) {BC};
\node[anchor=east,font=\scriptsize] at (axis cs:11.2,14.97) {AB};
\draw[<->,thick] (axis cs:11.7,23) -- (axis cs:22.3,23)
node[midway,above,font=\scriptsize,align=center] {Identical Power\\Despite Population Gap};
\draw[dashed] (axis cs:11.7,15.5) -- (axis cs:11.7,22.5);
\draw[dashed] (axis cs:22.3,15.5) -- (axis cs:22.3,22.5);
\node[anchor=west,font=\scriptsize] at (axis cs:4.5,4) {Small provinces ($\approx 0$)};
\end{axis}
\end{tikzpicture}
\caption{Population share versus Banzhaf power in the population-only system. Flat horizontal patterns indicate distinct demographic weights yielding identical power.}
\label{fig:scatter}
\end{figure}

\subsubsection*{Finding 2: Dilution of Hegemony}
The transition from a single-threshold to a dual-threshold system fundamentally reshapes the strategic landscape at the top. Under the population-only model, power is highly concentrated: Ontario emerges as a near-dominant player with a Banzhaf index of approximately $0.50$. This means Ontario is critical in roughly half of all winning coalitions.

However, the 7/50 rule imposes a cardinality constraint. Despite its massive $38.8\%$ population share, Ontario contributes only one discrete unit toward the provincial threshold. As a result, its Banzhaf index drops to $0.1463$. Ontario must now secure the support of at least six additional provinces regardless of their population size. Introducing a cardinality constraint therefore reduces the dominance of large weights and forces coalition formation into a higher-dimensional constraint space.

\subsubsection*{Finding 3: Resurrection of Small Provinces and the Power Plateau}
The most striking transformation occurs among the smallest provinces. In the population-only system, Prince Edward Island acts almost like a dummy player, being critical only 2 times out of 1023 possible coalitions. Under the 7/50 rule, both Prince Edward Island and Newfoundland and Labrador rise to a Banzhaf index of $0.0894$, a dramatic increase that makes them critical in 66 coalitions.

\begin{figure}[!t]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=0.98\columnwidth,
    height=0.78\columnwidth,
    symbolic x coords={Pop Only Rule,7/50 Rule},
    xtick=data,
    ymin=0,
    ymax=0.55,
    ylabel={Banzhaf Power Index},
    enlarge x limits=0.2,
    grid=major,
    grid style={dashed},
    legend pos=north east,
    legend cell align={left},
    tick label style={font=\scriptsize},
    label style={font=\scriptsize},
    legend style={font=\scriptsize},
    clip=false
]
\addplot[thick,mark=square*,mark size=2.5pt] coordinates {(Pop Only Rule,0.5000) (7/50 Rule,0.1463)};
\addlegendentry{Ontario (ON)}
\addplot[thick,mark=triangle*,mark size=3pt] coordinates {(Pop Only Rule,0.1497) (7/50 Rule,0.1057)};
\addlegendentry{QC, BC, AB}
\addplot[thick,mark=*,mark size=2.5pt] coordinates {(Pop Only Rule,0.0025) (7/50 Rule,0.0894)};
\addlegendentry{NL, PE}
\addplot[thick,mark=diamond*,mark size=3pt] coordinates {(Pop Only Rule,0.0127) (7/50 Rule,0.0894)};
\addlegendentry{MB, SK, NS}
\node[
    draw,
    fill=white,
    rounded corners=1pt,
    inner sep=1.5pt,
    font=\scriptsize,
    align=center,
    anchor=west
] (ppLabel) at (rel axis cs:0.58,0.34) {Power\\Plateau};
\draw[->,semithick] (ppLabel.east) to[out=0,in=180] (axis cs:7/50 Rule,0.0894);

\node[
    draw,
    fill=white,
    rounded corners=1pt,
    inner sep=1.5pt,
    font=\scriptsize,
    align=center,
    anchor=west
] (dhLabel) at (rel axis cs:0.50,0.22) {Diluted\\Hegemony};
\draw[->,semithick] (dhLabel.east) to[out=0,in=180] (axis cs:7/50 Rule,0.1463);
\end{axis}
\end{tikzpicture}
\caption{Structural impact of the 7/50 rule: reduced dominance for large provinces and increased influence for smaller provinces.}
\label{fig:trend}
\end{figure}

This dramatic shift is driven by the combinatorial structure of coalition formation. Because the $50\%$ population threshold is often satisfied by combinations of larger provinces, the provincial threshold becomes the binding constraint. When coalitions reach six provinces with sufficient population, the addition of a mathematically small seventh province becomes decisive. This makes several smaller provinces functionally interchangeable as pivotal members.

\begin{proposition}[Power Plateau Phenomenon]
In a dual-threshold voting system, if the population constraint is frequently satisfied independently, then smaller players become interchangeable in satisfying the cardinality constraint, leading to identical Banzhaf indices.
\end{proposition}

A further consequence is a power convergence phenomenon. Our results show that the Banzhaf indices of the six smaller provinces from Manitoba down to Prince Edward Island converge to the same value of $0.0894$. This power plateau indicates that differences in population size no longer generate differences in voting power. The rule effectively compresses variation, treating smaller provinces as interchangeable units required to satisfy the provincial quota rather than as weighted demographic actors.

\section{Limitations}
While the Banzhaf Power Index provides a clear combinatorial measure of voting power, it has several important limitations.

First, it assumes a uniform distribution over all coalitions, treating each subset as equally likely. This is an external modeling assumption and not a consequence of the institutional system itself.

Second, the index is based on binary pivotality: a player is either critical or not. As a result, microscopically small changes in weights or thresholds can sometimes produce disproportionately large changes in measured power.

Third, the model is static and combinatorial. Unlike permutation-based measures such as the Shapley--Shubik index, it does not incorporate ordering or temporal coalition formation.

Fourth, in multi-threshold systems such as the Canadian 7/50 rule, non-additive constraints sever any direct relationship between weight and power. Standard proportional intuitions can therefore fail dramatically.

Finally, the normalized index is inherently relative because each player's fractional power depends on the total number of swings across all players. A change affecting one player therefore changes the full distribution.

\section{Conclusion}
This project demonstrates that voting power in weighted systems is fundamentally a combinatorial phenomenon rather than a direct function of nominal voting weights. Through the Banzhaf Power Index, we showed that a player's influence depends on the frequency with which that player is pivotal across all possible coalitions.

In the theoretical game $[10:7,6,4,1]$, players with unequal weights can possess identical power, while a smaller player can become completely irrelevant. Extending the framework to the Canadian Constitutional Amending Formula shows that introducing a cardinality constraint fundamentally alters the combinatorial structure of winning coalitions. Large provinces lose dominance because they need broader coalition support, while small provinces gain influence by helping coalitions satisfy the provincial threshold.

Most notably, the emergence of a power plateau among smaller provinces demonstrates that, under binding multi-threshold constraints, voting power can converge across players despite population differences. More broadly, the analysis shows that voting systems must be evaluated through their mathematical structure. The Banzhaf framework provides a rigorous way to reveal hidden imbalances that are not visible from voting weights alone.

\begin{thebibliography}{00}

\bibitem{banzhaf}
J.~F.~Banzhaf III, ``Weighted voting doesn't work: A mathematical analysis,'' \emph{Rutgers Law Review}, vol.~19, no.~2, pp.~317--343, 1965.

\bibitem{felsenthal}
D.~S.~Felsenthal and M.~Machover, \emph{The Measurement of Voting Power: Theory and Practice, Problems and Paradoxes}. Cheltenham, U.K.: Edward Elgar, 1998.

\bibitem{taylor}
A.~D.~Taylor and W.~S.~Zwicker, \emph{Simple Games: Desirability Relations, Trading, Pseudoweightings}. Princeton, NJ, USA: Princeton Univ. Press, 1999.

\end{thebibliography}

\end{document}
